---
title: 'Reproducible Research: Peer Assessment 1'
author: "Borja Perez"
date: "27 of April, 2019"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

The following code is for loading the data, available in the activity.csv file:

```{r cars, echo = FALSE}
data <- read.csv(file="activity.csv", header=TRUE, sep=",")
```

## What is mean total number of steps taken per day?
The total steps taken per day is calculated with the following code. It also outputs a histogram of this measurement:

```{r }
days <- names(split(data, data$date))
res <- sapply(days, function(x){sum(data[data$date == x, 1])})
plot(as.Date(days), res, main="Histogram of steps/day", 
     xlab="Date (October to November 2012)",  type="h", lwd=6, col="red")
res

```

## What is the average daily activity pattern?
The following code makes a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

For it, firstly calculates the mean of every interval during each day ignoring NAs
```{r}
intervalSplit <- split(data$steps, data$interval)
averageStepsPerInterval <- sapply(intervalSplit, mean, na.rm=TRUE)
```
And then plots the time series graph
```{r}
plot(unique(data$interval), averageStepsPerInterval, type="l",
      main="Average num steps per interval across all days", 
      xlab="Interval", ylab="Average number of steps", 
      lwd=3, col="red")
abline(v = data[which.max(averageStepsPerInterval),3])
```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
The interval will be:
```{r}
data[which.max(averageStepsPerInterval),3]

```
## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as \color{red}{\verb|NA|}NA). The presence of missing days may introduce bias into some calculations or summaries of the data.
Calculate and report the total number of missing values in the dataset:
```{r}
sum(is.na(data$steps))
```
Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc. 
In this step, I will replace all the NA`s with the Mean value of that interval, and we will create a new dataset that is equal to the original dataset but with the missing data filled in.


```{r}
segment <- data[is.na(data$steps),] 
for(i in 1:dim(segment)[1]){
     for(j in 1:length(averageStepsPerInterval)){
         if(as.character(segment$interval[i]) == names(averageStepsPerInterval[j])){
             segment$steps[i] <- averageStepsPerInterval[j]
          }
    }
}
dataWithoutNAs <- data
dataWithoutNAs[rownames(segment),1] <- segment$steps
```
Create a histogram comparing the previous and the new one:
```{r }
par(mfrow=c(1,2))
plot(as.Date(days), res, main="Histogram of steps/day", 
     xlab="Date (October to November 2012)",  type="h", lwd=6, col="red")
res2 <- sapply(days, function(x){sum(dataWithoutNAs[dataWithoutNAs$date == x, 1])})

plot(as.Date(days), res2, main="Histogram of steps/day", 
     xlab="Date (October to November 2012)",  type="h", lwd=6, col="blue")


```
## Are there differences in activity patterns between weekdays and weekends?
In order to know if there are differences, we will split the dataset in two datasets: first one weeek days, second one weekend days:

```{r }
date <- strptime(dataWithoutNAs$date, "%Y-%m-%d")
dataWithoutNAs$weekDay <- date$wday
filtered <- split(dataWithoutNAs, dataWithoutNAs$weekDay==6 | dataWithoutNAs == 7)
weekDays <- filtered[[1]]
weekend <- filtered[[2]]
```
With this, we will look for differences for both dataframes. We will make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r }

intervalSplitWeekdays <- split(weekDays$steps, weekDays$interval)
averageStepsPerIntervalWeekdays <- sapply(intervalSplitWeekdays, mean, na.rm=TRUE)

intervalSplitWeekends <- split(weekend$steps, weekend$interval)
averageStepsPerIntervalWeekend <- sapply(intervalSplitWeekends, mean, na.rm=TRUE)

par(mfrow=c(1,2))
plot(unique(weekDays$interval), averageStepsPerIntervalWeekdays, type="l",
      main="Average steps per interval during week days/weekends", 
      xlab="Interval", ylab="Average number of steps weekdays", 
      lwd=2, col="red", ylim = c(0, 250))

plot(unique(weekend$interval), averageStepsPerIntervalWeekend, type="l",
      xlab="Interval", ylab="Average number of steps weekends", 
      lwd=2, col="blue", ylim = c(0, 250))

```

